---
title: 'Inferential Analysis of Tooth Growth Data'
author: "Hamed Bastan-Hagh"
date: "22 July 2016"
output: pdf_document
---

## Synopsis

This is an analysis of the ToothGrowth data in R. This records the length 
of odontoblasts (cells responsible for tooth growth) in 60 guinea pigs. Each 
animal received one of three dose levels of vitamin C (0.5, 1, and 2 mg/day) by 
one of two delivery methods: orange juice, coded as OJ; or ascorbic acid, a 
form of vitamin C, coded as VC.

This analysis concludes that there is a statistically significant effect of 
dose amount on tooth growth, but no such effect for supplement type.

## 1. Load Data and Perform Basic Exploratory Data Analysis

``` {r Loading the Data}
# Load required libraries
library(datasets)
library(dplyr)
library(ggplot2)
# Assign ToothGrowth data to variable df in 'tibble' form for easier printing
df <- tbl_df(ToothGrowth)
```

With the data loaded we will do some basic exploratory analysis.

``` {r Exploratory Data Analysis 1}
# Boxplot of the values separated by supplement type
ggplot(df, aes(supp, len)) +
    geom_boxplot(aes(fill = supp), notch = TRUE) +
    labs(title = "Effect of Supplement Type on Tooth Growth",
         x = "Supplement Type",
         y = "Length of Odontoblasts")
```

The notches for the two groups do not overlap, which is evidence that 
the median value for Orange Juice is higher than Ascorbic Acid. We will 
investigate this further in section 3.

``` {r Exploratory Data Analysis 2}
# Boxplot of the values separated by dose amount
ggplot(df, aes(factor(dose), len)) +
    geom_boxplot(aes(fill = dose)) +
    labs(title = "Effect of Dose On Tooth Growth",
         x = "Dose in mg/day",
         y = "Length of Odontoblasts")
```

This indicates a positive correlation between dose amount and length, which we 
will investigate further in section 3.

## 2. Summary of the Data

Here are some basic summaries for the data set.

``` {r Data Summary 1}
#Print put basic summary of our data
summary(df)
# Print out levels for supplement type 'supp'
levels(df$supp)
# Print out unique values for dose amount 'dose'
unique(df$dose)
```

Now we can see a summary of the data grouped by dose amount.

``` {r Data Summary 2}
df %>% 
    group_by(dose) %>% 
        summarise(len = mean(len))
```

And finally a summary of the data grouped by supplement type.

``` {r Data Summary 3}
df %>% 
    group_by(supp) %>% 
        summarise(len = mean(len))
```

## 3. Compare tooth growth by "supp" and "dose"

### 3a. Supplement Type

Firstly we will explore the effect of supplement type on tooth growth with a 
t-test. Our null hypothesis is:

`H_0: mu = mu_0`

The alternative hypothesis:

`H_a: mu != mu_0`

``` {r Testing Effect of Supplement Type on Tooth Growth}
# Use function t.test on the dataset
t.test(len ~ supp, df)
```

Our p-value is 0.06, which is slightly higher than our threshold for 95% 
confidence. Therefore we **fail to reject the null hypothesis**, and cannot 
conclude that supplement type has an effect on tooth growth.

### 3b. Dose Amount

Now we will test the effect of dose amount on tooth growth in our data. We will 
test the ToothGrowth data in three sections based on dose amount: testing 0.5 
against 1.0; 0.5 against 2.0; and 1.0 against 2.0. For each of these our null 
hypothesis is that:

`H_0: mu = mu_0`

Alternative hypothesis:

`H_a: mu != mu_0`

``` {r Testing Effect of Dose Amount on Tooth Growth 1}
# Testing 0.5 against 1.0 by removing those rows where dose = 2.0
t.test(len ~ dose, df[df$dose != 2, ])
```

``` {r Testing Effect of Dose Amount on Tooth Growth 2}
# Testing 0.5 against 2.0 by removing those rows where dose = 1.0
t.test(len ~ dose, df[df$dose != 1, ])
```

``` {r Testing Effect of Dose Amount on Tooth Growth 3}
# Testing 1.0 against 2.0 by removing those rows where dose = 0.5
t.test(len ~ dose, df[df$dose != 0.5, ])
```

P-values for each of these tests is low enough that we can **reject the null 
hypothesis**. We can conclude that dose amount does have an impact on tooth 
growth.

## 4. Conclusions

Based on our analysis we can conclude that:

- Supplement type does not have a statistically significant effect on tooth 
growth in our data.
- Dose amount does have a statistically significant effect on tooth growth in 
our data.