---
title: 'Statistical Inference: Final Project; Part 1'
author: "Hamed Bastan-Hagh"
date: "20 July 2016"
output: pdf_document
---

## The Exponential Distribution and the Central Limit Theorem

### Overview
In this project I will investigate the exponential distribution in R and 
compare it with the Central Limit Theorem. The exponential distribution can be 
simulated in R with `rexp(n, lambda)` where lambda is the rate parameter. The 
mean of the exponential distribution is 1/lambda and the standard deviation is 
also 1/lambda. I will set lambda = 0.2 for all of the simulations. I will 
investigate the distribution of averages of 40 exponentials. I will illustrate 
via 1,000 simulations and associated explanatory text the properties of the 
distribution of the mean of 40 exponentials.

We will see that the distributions of the sample means and variance are as we 
would expect from the Central Limit Theorem in our simulations.

### Question 1:  
**Compare the sample mean to the theoretical mean of the distribution.**

To answer this we need 1,000 samples with size n = 40 from the exponential 
distribution. We can simulate this using the `rexp` function, and then plot the 
means of each of those samples in a histogram.

``` {r First simulation}
# Load stats library
library(stats)
# Set n to 40, number of sims to 1,000 and lambda to 0.2 as directed
n <- 40
sims <- 1000
lambda <- 0.2
# Set seed to make this reproducible
set.seed(1981)
# We will generate 40,000 random values from the exponential distribution 
# using rexp. These go into a 1000 x 40 matrix: each of the 1,000 rows is one 
# sample with size 40.
data <- matrix(rexp(n = n * sims, rate = lambda), sims)
# Calculate the means for each of these samples
samp.mean <- apply(data, 1, mean)
# Show histogram of the distribution, with vertical lines at the sample mean
# and the theoretical mean
library(ggplot2)
xbar <- mean(samp.mean)
mu <- 1 / lambda
ggplot(data = data.frame(Means = samp.mean), aes(x = Means)) +
    geom_histogram(colour = "black", fill = "lightgrey", binwidth = 0.2) +
    geom_vline(xintercept = xbar, colour = "red") +
    geom_vline(xintercept = mu, colour = "blue") +
    labs(title = "Distribution of Sample Means From the Exponential 
         Distribution",
         x = "Sample Means",
         y = "Count")
```

NB. The red vertical line shows the sample mean, while the blue line is the 
theoretical mean. They are very close to each other so you may find it 
difficult to see the red line.

#### Sample Mean
The sample mean is `r round(xbar, 3)`

#### Theoretical Mean
The theoretical mean is `1 / lambda`, which is `r mu`.

#### Comparison
The difference between the sample and theoretical means is only 
`r abs(xbar - mu)`, so we can see that the sample mean is a close approximation 
to the theoretical mean.

### Question 2: 
**Compare the variance of the sample to the theoretical variance of the 
distribution.**

The variance of our distribution of sample means is 
**`r round(var(samp.mean), 2)`.**

The theoretical variance of an exponential distribution is `1 / lambda` = 
**`r 1 / lambda`**. So the expected variance of our sample means is 
`(1/n) * (1/lambda^2)` = **`r (1/n) * (1/lambda^2)`**. **The difference between 
these is `r abs(var(samp.mean) - ((1/n) * (1/lambda^2)))`**. We can see that 
the variance of our sample means is again very close to the theoretical figure, 
as we would expect from the Central Limit Theorem.

### Question 3:
**Show that the distribution is approximately normal.**

We can demonstrate that the distribution is approximately normal by overlaying 
the curve for the normal distribution on the histogram of the distribution of 
the sample means.

``` {r Plotting Normal Distribution With Sampled Means}
# Assign the variance and standard deviation of the sample to variables
samp.var <- (1/n) * (1/lambda^2)
samp.sd <- sqrt(samp.var)
# Plot the distribution of the sample means with mean and normal curve 
# overlaid
ggplot(data = data.frame(Means = samp.mean), aes(x = Means)) +
    geom_histogram(colour = "black", 
                   fill = "lightgrey", 
                   binwidth = 0.1,
                   aes(y = ..density..)) +
    geom_vline(xintercept = mu, colour = "blue") +
    stat_function(fun = dnorm, 
                  color = "red",
                  args = list(mean = mu,
                              sd = samp.sd)) +
    labs(title = "Distribution of Sample Means From the Exponential 
         Distribution \n(Overlaid with Normal Distribution)",
         x = "Sample Means",
         y = "Density")
```

**This shows is that the distribution of the sample means is approximately 
normal.**